Algebraic Principles for Concurrency Verification Tools

### 2013-10-29 Paper/paper.tex
Started re-working paper

\DeclareFontFamily{U}{wncy}{}
\DeclareFontShape{U}{wncy}{m}{n}{<->wncyss10}{}
\DeclareSymbolFont{mcy}{U}{wncy}{m}{n}
\DeclareMathSymbol{\sh}{\mathord}{mcy}{"78} 
\DeclareMathOperator{\sha}{\sh}
\DeclareMathOperator{\tr}{Tr}
\DeclareMathOperator{\fin}{finite}
\DeclareMathOperator{\rely}{rely}
\DeclareMathOperator{\guar}{guar}
\DeclareMathOperator{\wlp}{wlp}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Algebra for Rely-Guarantee}
\label{sec:RG}

Define a rely-guarantee algebra as a structure
$(K,RG,+,\sqcap,\cdot,\|,^\star,0,1)$, where $(K,+,\sqcap)$ is a
lattice, $(K,+,\cdot,\|,0,1)$ is a weak trioid and
$(K,+,\cdot,^\star,0,1)$ is a Kleene algebra. $RG$ is a distinguished
subset of relys and guarantees which satisfy the following axioms
\begin{align}
r\|r &\le r, \label{rg1}\\
r &\le r\|r', \label{rg2}\\
r\|xy &= (r\|x)(r\|y), \label{rg3}\\
r\|x^+ &\le (r\|x)^+ \label{rg4}.
\end{align}
By convention, we use $r$ and $g$ to refer to elements of $RG$, and
$x,y,z$ for arbitrary elements of $K$. Some elementary consequences of these rules are as follows
\begin{align*}
1 &\le r,\\
r^\star = rr &= r = r\|r,\\
r\|x^+ &= (r\|x)^+.
\end{align*}

@@

Axiom (\ref{rg4}) can be derived from (\ref{rg3}) in all finite
models. This is because any finite $K$ is complete, and in a complete
setting fixpoint laws can be used to show $r\|x^+ \le (r\|x)^+$. Thus
it is impossible to construct a finite model demonstrating that
(\ref{rg4}) is independent from (\ref{rg1}) -- (\ref{rg2}).

@@

We can expand our rely-guarantee algebras to have residuals
axiomatised by Galois connections for both parallel and sequential
composition, i.e.
\begin{align}
xy \le z &\iff y \le x \rightarrow z, \label{res1}\\
xy \le z &\iff x \le z \leftarrow y, \label{res2}\\
x\|y \le z &\iff y \le x/z \label{res3}.
\end{align}
For sequential composition, (\ref{res1}) and (\ref{res2}) essentially
provide a link between weakest-precondition style semantics and hoare
logic. As mentioned in Section \ref{sec:KA}, the triple $\{p\} c
\{q\}$ can be encoded as $pc \le q$, while $q \leftarrow c$ represents
$wlp(c,q)$. Equation (\ref{res2}) can thus be written as $\{p\} c
\{q\} \iff p \le wlp(c,q)$. The parallel residual $x/z$ can be
understood similarly as follows; it is the weakest program such that
when placed in parallel with $x$, the composition behaves as $z$.

By using residuals quintuples can be encoded in the following way,
which is equivalent to the encoding in Equation (\ref{quin}).
\begin{align}
r, g \vdash \{p\} c \{q\} \iff c \le r/(p \rightarrow q) \sqcap g \label{refine}.
\end{align}
This encoding allows us to think in terms of program refinement, as
$r/(p \rightarrow q) \sqcap g$ defines the weakest program that when
placed in parallel with interference from $r$, and guaranteeing
interference at most $g$, goes from $p$ to $q$---which is essentially
a generic specification for a concurrent program.

@@

While the algebra in section \ref{sec:RG} is adequate for deriving the
standard inference rules, the equality is too strong to capture many
interesting statements about concurrent programs. Consider the
congruence rule for parallel composition, which is inherent in the
algebraic approach:
\begin{align*}
x = y \implies x\|z = y\|z.
\end{align*}
This can be read as follows; if $x$ and $y$ are equal, then they must
be equal under all possible interference from any arbitrary $z$. At
first glance this might seem to preclude any reasoning about
interference in a fine-grained way using only algebra. This is not the
case, but breaking such inherent compositionality in just the right way to
capture interesting properties of interference requires some extra work.

The most obvious way we can acheive this is by simply expanding our
rely-guarantee algebra with an arbitrary function $f$. And redefining our quintuples
as,
\begin{align*}
r, g \vdash \{p\} c \{q\} \iff f(p(r\|c)) \le f(q) \land c \le g.
\end{align*}

 but we can work around this by
using an \emph{interior operator}. In a rely-guarantee algebra an
interior operator is any function $f : K \to K$ which satisfies
\begin{align*}
f(f(x)) &= f(x) \tag{$f$ is idempotent}\\
x \le y \implies f(x) &\le f(y) \tag{$f$ is isotone}\\
f(x) &\le x \tag{$f$ is coextensive}
\end{align*}

We extend our rely-guarantee algebra to
$(K,RG,+,\cdot,\|,^\star,[\--],0,1)$ by adding a single interior
operator $[\--]$, which additionally satisfies
\begin{align}
[xy] &= [x][y], \label{rgimult}\\
[x + y] &= [x] + [y],\\
[x \sqcap y] &= [x] \sqcap [y],\\
[x^\star] &= [x]^\star,\\
[0] &= 0 \label{rgi0}
\end{align}
Intuitively $[x]$ can be thought of as the \emph{execution} of a
program $x$. While $x$ contains all information about what the process
or program \emph{could} do, $[x]$ only contains information about what
it \emph{actually} does. Note that (\ref{rgimult}) -- (\ref{rgi0}) do
not mention parallel composition, such that in general $[x] = [y]
\centernot\implies [x\|z] = [y\|z]$. We amend our jones-quintuples to
involve this operator as shown below.
\begin{align*}
r,g \vdash \{p\} c \{q\} \iff p[r\|c] \le q \land c \le g.
\end{align*}
All the rules in Figure \ref{fig:rgrules} can be derived with this
encoding, and moreover their proofs remain the same, mutatis
mutandis.

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Algebra for Separation Logic}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Finite Language Model}
\label{sec:Model}

@@

As usual, the product of two languages X and Y over an alphabet
$\Sigma$ is defined as $XY = \{x\frown y.\; x \in X \land y \in Y\}$.
The language product is associative, and distributes over arbitrary
joins from both the left and the right, i.e.
\begin{align*}
X(\bigcup Y) = \bigcup\{XZ.\; Z \in Y\} \qquad \text{and} \qquad (\bigcup Y)X = \bigcup\{ZX.\; Z \in Y\}.
\end{align*}
The shuffle $\sha$ on a pair
of finite words is defined inductively as
\begin{align*}
\epsilon \sha s = \{s\}, \qquad s \sha \epsilon = \{s\},\\
as \sha bt = a(s \sha bt) \cup b(as \sha t),
\end{align*}
which is then lifted to languages X and Y as the shuffle product,
\begin{align*}
X \sha Y = \bigcup\{xs \sha ys.\; x \in X \land x \in Y\}.
\end{align*}
The shuffle product is associative, commutative, and distributes over
arbitrary joins. Both products share the same unit, $\{\epsilon\}$ and
zero, $\emptyset$. Proving properties of shuffle is suprisingly
tricky (especially if one considers infinite words). For a in-depth
treatment of the shuffle product see~\cite{shufflethings}. The
above properties are enough to show that $(\mathcal{P}(\Sigma)^*),
\cup, \cdot,\sha, \emptyset, \{\epsilon\})$ forms a weak trioid.

The rely-guarantee elements in this model are sets containing all the
words which can be built from some set of actions in $\Sigma$. We can
consider a function $\rely : \mathcal{P}(\Sigma) \to
\mathcal{P}(\Sigma)^*$ which takes a set of actions and returns all
the words that can be built from them. This function is an upper
adjoint for $\guar$, the function which returns all the letters in a
language.
\begin{align*}
\guar(X) \le Y \iff X \le \rely(Y)
\end{align*}

Properties (\ref{rg1}) and (\ref{rg2}) from Section \ref{sec:RG} can
mostly be derived from basic properties of $\rely$. Namely,
\begin{align*}
\rely(X \cup Y) &= \rely(X) \| \rely(Y),\\
\qquad \rely(X \cap Y) &= \rely(X) \cap \rely(Y).
\end{align*}
The proofs of (\ref{rg3}) and (\ref{rg4}) are somewhat more involved,
but ultimately rather mechanical, so we do not show them here. We
define the set $RG$ of rely-guarantee elements as the range of the
$\rely$ function and therefore are able to show that $(\mathcal{P}(\Sigma)^*), RG, \cup,
\cdot,\sha, \emptyset, \{\epsilon\})$ is a rely-guarantee algebra.

To make our model more concrete we consider an alphabet comprising
relations ($R,S,T,\dots$) over a state space $\Sigma$. A single
relation represents a single atomic action of a program. Assuming
$\Sigma = \{\sigma_1,\sigma_2,\sigma_3\}$, an example of a word in
$(\Sigma^2)^*$ is shown below.

\begin{center}
\begin{tikzpicture}[x=1cm,auto]
  \node (center) {};
  \node [elem] (s12) {$\sigma_2$};
  \node [elem, above of=s12] (s11) {$\sigma_1$};
  \node [elem, below of=s12] (s13) {$\sigma_3$};
  \node [right of=s13, node distance=0.75cm] (r1) {};
  \node [below of=r1, node distance=0.5cm] (t4) {R};

  \node [elem, right of=s12, node distance=1.5cm] (s22) {$\sigma_2$};
  \node [elem, above of=s22] (s21) {$\sigma_1$};
  \node [elem, below of=s22] (s23) {$\sigma_3$};
  \node [right of=s23, node distance=0.75cm] (r2) {};
  \node [below of=r2, node distance=0.5cm] (t4) {S};

  \node [elem, right of=s22, node distance=1.5cm] (s32) {$\sigma_2$};
  \node [elem, above of=s32] (s31) {$\sigma_1$};
  \node [elem, below of=s32] (s33) {$\sigma_3$};
  \node [right of=s33, node distance=0.75cm] (r3) {};
  \node [below of=r3, node distance=0.5cm] (t4) {T};

  \node [elem, right of=s32, node distance=1.5cm] (s42) {$\sigma_2$};
  \node [elem, above of=s42] (s41) {$\sigma_1$};
  \node [elem, below of=s42] (s43) {$\sigma_3$};
  \node [right of=s43, node distance=0.75cm] (r4) {};
  \node [below of=r4, node distance=0.5cm] (t4) {U};

  \node [elem, right of=s42, node distance=1.5cm] (s52) {$\sigma_2$};
  \node [elem, above of=s52] (s51) {$\sigma_1$};
  \node [elem, below of=s52] (s53) {$\sigma_3$};

  \path [line] (s11) -- (s21);
  \path [line] (s21) -- (s32);
  \path [line] (s32) -- (s43);
  \path [line] (s43) -- (s52);

  \path [rel] (s11) -- (s22);
  \path [rel] (s12) -- (s22);
  \path [rel] (s23) -- (s32);
  \path [rel] (s32) -- (s42);
  \path [rel] (s42) -- (s51);
  \path [rel] (s41) -- (s52);
\end{tikzpicture}
\end{center}

This word only has a single valid \emph{trace}, which is
$\sigma_1\rightarrow\sigma_1\rightarrow\sigma_2\rightarrow\sigma_3\rightarrow\sigma_2$.
Such state transition traces are commonly called \emph{Aczel
  traces}~\cite{Aczel}. Note that unlike much of the literature involving
Aczel-style traces for concurrency, we omit any kind of process or
environment labelling from our transitions. We define an operator
$\tr(X)$ which returns all such traces in a language $X$. $\tr$ is
both idempotent and isotone.

Note that we cannot use the shuffle operator on directly traces. For
example, imagine we shuffled the single transition $\sigma_2 \to
\sigma_3$ into the above word, we could end up with a word as below.

\begin{center}
\begin{tikzpicture}[x=1cm,auto]
  \node [elem] (s12) {$\sigma_2$};
  \node [elem, above of=s12] (s11) {$\sigma_1$};
  \node [elem, below of=s12] (s13) {$\sigma_3$};
  \node [right of=s13, node distance=0.75cm] (r1) {};
  \node [below of=r1, node distance=0.5cm] (t4) {R};

  \node [elem, right of=s12, node distance=1.5cm] (i2) {$\sigma_2$};
  \node [elem, above of=i2] (i1) {$\sigma_1$};
  \node [elem, below of=i2] (i3) {$\sigma_3$};
  \node [right of=i3, node distance=0.75cm] (ri) {};
  \node [below of=ri, node distance=0.5cm] (ti) {V};

  \node [elem, right of=i2, node distance=1.5cm] (s22) {$\sigma_2$};
  \node [elem, above of=s22] (s21) {$\sigma_1$};
  \node [elem, below of=s22] (s23) {$\sigma_3$};
  \node [right of=s23, node distance=0.75cm] (r2) {};
  \node [below of=r2, node distance=0.5cm] (t4) {S};

  \node [elem, right of=s22, node distance=1.5cm] (s32) {$\sigma_2$};
  \node [elem, above of=s32] (s31) {$\sigma_1$};
  \node [elem, below of=s32] (s33) {$\sigma_3$};
  \node [right of=s33, node distance=0.75cm] (r3) {};
  \node [below of=r3, node distance=0.5cm] (t4) {T};

  \node [elem, right of=s32, node distance=1.5cm] (s42) {$\sigma_2$};
  \node [elem, above of=s42] (s41) {$\sigma_1$};
  \node [elem, below of=s42] (s43) {$\sigma_3$};
  \node [right of=s43, node distance=0.75cm] (r4) {};
  \node [below of=r4, node distance=0.5cm] (t4) {U};

  \node [elem, right of=s42, node distance=1.5cm] (s52) {$\sigma_2$};
  \node [elem, above of=s52] (s51) {$\sigma_1$};
  \node [elem, below of=s52] (s53) {$\sigma_3$};

  \path [rel] (s11) -- (i1);
  \path [rel] (s21) -- (s32);
  \path [line] (s32) -- (s43);
  \path [line] (s43) -- (s52);
  \path [line] (i2) -- (s23);

  \path [rel] (s11) -- (i2);
  \path [line] (s12) -- (i2);
  \path [line] (s23) -- (s32);
  \path [line] (s32) -- (s42);
  \path [line] (s42) -- (s51);
  \path [rel] (s41) -- (s52);
\end{tikzpicture}
\end{center}

By performing this shuffle action, we no longer have a trace from
$\sigma_1$ to $\sigma_2$, but instead two traces from $\sigma_2$ to
$\sigma_2$ and $\sigma_3$. This subtle interplay between shuffle and
$\tr$ mimics how Aczel-traces are typically generated from an
operational semantics in much existing rely-guarantee literature. This
operational approach is not desireable from an algebraic perspective
though, as we would rather see programs as mathematical objects in
their own right---a more denotational approach.

@@

Next, we define the following language for a word $w = R_0R_1\dots
R_n$. The language $w^\udagger$ contains all strings $S_0S_1\dots S_n$
such that for all $m \le n$, $S_m \subseteq R_m$. Effectively this
gives us all the words which are less that $w$ pointwise. Again this
operation is lifted to languages in the same way as before. We define
the set of \emph{$smp$-closed} languages by combining both these
closure operators into a single closure operator, $X^\ddagger$, which
satisfies the properties
\begin{align}
(X^\ddagger \sha Y)^\ddagger &= (X \sha Y)^\ddagger,\\
(X^\ddagger Y)^\ddagger = (&XY)^\ddagger = (XY^\ddagger)^\ddagger,\\
(X \cup Y)^\ddagger &= (X^\ddagger \cup Y^\ddagger),\\
(X^\ddagger \cap Y^\ddagger)^\ddagger &= (X^\ddagger \cap Y^\ddagger),\\
\tr(X^\ddagger) &\subseteq X^\ddagger. \label{trcoext}
\end{align}
These five properties essentially allow us to `lift' all our operators
from above to act on $smp$-closed languages in such a way that they
still form a weak trioid. Henceforth, we will assume all our languages are
$smp$-closed, and all operators have been appropriately lifted.

Equation (\ref{trcoext}) shows that $\tr$ is now an interior operator
on the set of $smp$-closed languages.

We now consider how tests are represented in this model, we define
$\langle R\rangle$ as the singleton language containing just the
relation $R$. Tests are defined as traces of the form $\langle
R\rangle^\ddagger$ where $R \subseteq I$. We denote the set of all
tests as $B$, and let it be ranged over by $P,Q,S$. From the
definition of $smp$-closure $\langle I\rangle^\ddagger =
\{\epsilon\}^\ddagger$, thus the greatest test is $1$, while the least
test is $\langle \emptyset\rangle^\ddagger$, $0$. As such our tests
form a boolean algebra between $1$ and $0$, much they would in
Kozen's Kleene algebra with tests.

\begin{center}
\begin{tikzpicture}[x=1cm,auto]
  \node (center) {};
  \node [elem] (s12) {$\sigma_2$};
  \node [elem, above of=s12] (s11) {$\sigma_1$};
  \node [elem, below of=s12] (s13) {$\sigma_3$};
  \node [right of=s13, node distance=0.75cm] (r1) {};
  \node [below of=r1, node distance=0.5cm] (t4) {$R$};

  \node [elem, right of=s12, node distance=1.5cm] (s22) {$\sigma_2$};
  \node [elem, above of=s22] (s21) {$\sigma_1$};
  \node [elem, below of=s22] (s23) {$\sigma_3$};
  \node [right of=s23, node distance=0.75cm] (r2) {};
  \node [below of=r2, node distance=0.5cm] (t4) {$R$};

  \node [elem, right of=s22, node distance=1.5cm] (s32) {$\sigma_2$};
  \node [elem, above of=s32] (s31) {$\sigma_1$};
  \node [elem, below of=s32] (s33) {$\sigma_3$};

  \node [elem, right of=s32, node distance=1.5cm] (s42) {$\sigma_2$};
  \node [elem, above of=s42] (s41) {$\sigma_1$};
  \node [elem, below of=s42] (s43) {$\sigma_3$};
  \node [right of=s43, node distance=0.75cm] (r4) {};
  \node [below of=r4, node distance=0.5cm] (t4) {$R$};

  \node [elem, right of=s42, node distance=1.5cm] (s52) {$\sigma_2$};
  \node [elem, above of=s52] (s51) {$\sigma_1$};
  \node [elem, below of=s52] (s53) {$\sigma_3$};
  \node [right of=s53, node distance=0.75cm] (r5) {};
  \node [below of=r5, node distance=0.5cm] (t5) {$\lnot R$};

  \node [elem, right of=s52, node distance=1.5cm] (s62) {$\sigma_2$};
  \node [elem, above of=s62] (s61) {$\sigma_1$};
  \node [elem, below of=s62] (s63) {$\sigma_3$};

  \path [line] (s11) -- (s21);
  \path [line] (s12) -- (s22);
  \path [line] (s21) -- (s31);
  \path [line] (s22) -- (s32);

  \path [rel] (s42) -- (s52);
  \path [rel] (s41) -- (s51);
  \path [rel] (s53) -- (s63);
\end{tikzpicture}
\end{center}

A key difference however is that tests only behave as tests when under
$\tr$. This is illustrated above---$\tr(\langle R\rangle^\ddagger\langle
R\rangle^\ddagger) = \tr(\langle R\rangle^\ddagger)$ while $\tr(\langle
R\rangle^\ddagger\langle \lnot R\rangle^\ddagger) = 0$.

>>>>

\DeclareFontFamily{U}{wncy}{}
\DeclareFontShape{U}{wncy}{m}{n}{<->wncyss10}{}
\DeclareSymbolFont{mcy}{U}{wncy}{m}{n}
\DeclareMathSymbol{\sh}{\mathord}{mcy}{"78} 
\DeclareMathOperator{\sha}{\sh}
\DeclareMathOperator{\tr}{Tr}
\DeclareMathOperator{\fin}{finite}
\DeclareMathOperator{\rely}{rely}
\DeclareMathOperator{\guar}{guar}
\DeclareMathOperator{\letters}{letters}
\DeclareMathOperator{\wlp}{wlp}

@@

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Algebra for Rely-Guarantee}
\label{sec:RG}

A rely-guarantee algebra is a structure
$(K,RG,+,\sqcap,\cdot,\|,^\star,0,1)$, where $(K,+,\sqcap)$ is a
distributive lattice, $(K,+,\cdot,\|,0,1)$ is a weak trioid and
$(K,+,\cdot,^\star,0,1)$ is a Kleene algebra. $RG$ is a distinguished
subset of relys and guarantees which satisfy the following axioms
\begin{align}
r\|r &\le r, \label{rg1}\\
r &\le r\|r', \label{rg2}\\
r\|xy &= (r\|x)(r\|y), \label{rg3}\\
r\|x^+ &\le (r\|x)^+ \label{rg4}.
\end{align}
By convention, we use $r$ and $g$ to refer to elements of $RG$, and
$x,y,z$ for arbitrary elements of $K$. Some elementary consequences of these rules are as follows
\begin{align*}
1 &\le r,\\
r^\star = rr &= r = r\|r,\\
r\|x^+ &= (r\|x)^+.
\end{align*}

@@

Axiom (\ref{rg4}) can be derived from (\ref{rg3}) in all finite
models. This is because any finite $K$ is complete, and in a complete
setting fixpoint laws can be used to show $r\|x^+ \le (r\|x)^+$. Thus
it is impossible to construct a finite model demonstrating that
(\ref{rg4}) is independent from (\ref{rg1}) -- (\ref{rg3}).

@@

If we expand our rely-guarantee algebras to have residuals axiomatised
by Galois connections for both parallel and sequential composition,
i.e.
\begin{align}
xy \le z &\iff y \le x \rightarrow z, \label{res1}\\
xy \le z &\iff x \le z \leftarrow y, \label{res2}\\
x\|y \le z &\iff y \le x/z \label{res3},
\end{align}
then for sequential composition, (\ref{res1}) and (\ref{res2}) provide
a link between weakest-precondition style semantics and hoare
logic. As mentioned in Section \ref{sec:KA}, the triple $\{p\} c
\{q\}$ can be encoded as $pc \le q$, while $q \leftarrow c$ represents
$wlp(c,q)$. Equation (\ref{res2}) can thus be written as $\{p\} c
\{q\} \iff p \le wlp(c,q)$. The parallel residual $x/z$ can be
understood similarly as follows; it is the weakest program such that
when placed in parallel with $x$, the composition behaves as $z$.

By using these residuals quintuples can be encoded in the following way,
which is equivalent to the encoding in Equation (\ref{quin}).
\begin{align}
r, g \vdash \{p\} c \{q\} \iff c \le r/(p \rightarrow q) \sqcap g \label{refine}.
\end{align}
This encoding allows us to think in terms of program refinement, as
$r/(p \rightarrow q) \sqcap g$ defines the weakest program that when
placed in parallel with interference from $r$, and guaranteeing
interference at most $g$, goes from $p$ to $q$---a generic
specification for a concurrent program.

@@

While the algebra in section \ref{sec:RG} is adequate for deriving the
standard inference rules, it's equality is too strong to capture many
interesting statements about concurrent programs. Consider the
congruence rule for parallel composition, which is inherent in the
algebraic approach:
\begin{align*}
x = y \implies x\|z = y\|z.
\end{align*}
This can be read as follows; if $x$ and $y$ are equal, then they must
be equal under all possible interference from any arbitrary $z$. At
first sight, this might seem to preclude any reasoning about
interference in a fine-grained way using purely algebra. This is not
the case, but breaking such inherent compositionality in just the
right way to capture interesting properties of concurrency requires
extra work.

The most obvious way we can acheive this is by expanding our
rely-guarantee algebra with an additional function $\pi$ and redefining
our quintuples as,
\begin{align*}
r, g \vdash \{p\} c \{q\} \iff \pi(p(r\|c)) \le \pi(q) \land c \le g.
\end{align*}
Since for any operator $\bullet$, it is not required that
\begin{align*}
\pi(x) = \pi(y) \implies \pi(x \bullet z) = \pi(y \bullet z),
\end{align*}
we can `break' compositionality in just the right way, provided we
choose the correct properties for $\pi$. For example, if we consider
the following seven axioms
\begin{align}
\pi(\pi(x)) &= \pi(x), \label{pi1}\\
x \le y \implies \pi(x) &\le \pi(y), \label{pi2}\\
\pi(x) &\le x, \label{pi3}\\
\pi(x^*) &\le \pi(\pi(x)^*), \label{pi4}\\
\pi(xy) &\le \pi(\pi(x)\pi(y)), \label{pi5}\\
\pi(x \sqcap y) &= \pi(x) \sqcap \pi(y). \label{pi6}
\end{align}

Axioms (\ref{pi1}) to (\ref{pi3}) show that $\pi$ is an interior
operator, while (\ref{pi4}) and (\ref{pi5}) allow us to introduce
$\pi$ as needed around sequential composition and finite iteration,
which allows us to substitute under those operators as required---we
only want to break compositionality for $\|$.

The above axioms indicate that we could define $\pi$ as a function
$\lambda x.\; x \sqcap c$ where $c$ satisfies
\begin{align}
x^* \sqcap c &\le (x \sqcap c)^* \sqcap c, \label{con1}\\
xy \sqcap c &\le (x \sqcap c)(y \sqcap c) \sqcap c. \label{con2}
\end{align}

TODO: Define algebra w/ c.

All the rules in Figure \ref{fig:rgrules} can be derived with this
algebra, and moreover their proofs remain the same, mutatis
mutandis.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Finite Language Model}
\label{sec:Model}

@@

We consider languages where the alphabet contains state pairs of the
form $(\sigma_1,\sigma_2) \in \Sigma \times \Sigma$. A word in such a
language is \emph{consistent} if every such pair in a word has the
same first state as the previous transition's second state. For
example, $(\sigma_1,\sigma_2)(\sigma_2,\sigma_3)$ is consistent, while
$(\sigma_1,\sigma_2)(\sigma_3,\sigma_3)$ is inconsistent. Sets of
consistent words are essentially \emph{Aczel
  traces}~\cite{boer_formal_1999}, but lack the usual process
labels. We denote the set of all consistent words by $C$.

As usual, the product of two languages X and Y is defined as
\begin{align*}
XY = \{x\frown y.\; x \in X \land y \in Y\}.
\end{align*}
The language product is associative, and distributes over arbitrary
joins from both the left and the right, i.e.
\begin{align*}
X(\bigcup Y) = \bigcup\{XZ.\; Z \in Y\} \qquad \text{and} \qquad (\bigcup Y)X = \bigcup\{ZX.\; Z \in Y\}.
\end{align*}
The shuffle $\sha$ on a pair
of finite words is defined inductively as
\begin{align*}
\epsilon \sha s = \{s\}, \qquad s \sha \epsilon = \{s\},\\
as \sha bt = a(s \sha bt) \cup b(as \sha t),
\end{align*}
which is then lifted to languages X and Y as the shuffle product,
\begin{align*}
X \sha Y = \bigcup\{x \sha y.\; x \in X \land x \in Y\}.
\end{align*}
The shuffle product is associative, commutative, and distributes over
arbitrary joins. Both products share the same unit, $\{\epsilon\}$ and
zero, $\emptyset$. Proving properties of shuffle is suprisingly
tricky (especially if one considers infinite words). For a in-depth
treatment of the shuffle product see~\cite{shufflethings}. The
above properties are enough to show that $(\mathcal{P}((\Sigma\times\Sigma)^*),
\cup, \cdot,\sha, \emptyset, \{\epsilon\})$ forms a weak trioid.

The rely-guarantee elements in this model are sets containing all the
words which can be built from some set of state pairs in
$\Sigma\times\Sigma$. We define a function $\langle R\rangle$ which
lifts a relation $R$ to a language containing length 1 words for each
pair in $R$. The set of rely-guaratee elements RG is then defined as
$\{r.\; \exists R. r = \langle R\rangle^*\}$. We can show that
$(\mathcal{P}((\Sigma\times\Sigma)^*), RG, \cup, \cdot,\sha,
\emptyset, \{\epsilon\}, C)$ is a rely-guarantee algebra.

To demonstrate how this model works, consider the graphical
representation of a language shown below.

\begin{center}
\begin{tikzpicture}[x=1cm,auto]
  \node (center) {};
  \node [elem] (s12) {$\sigma_2$};
  \node [elem, above of=s12] (s11) {$\sigma_1$};
  \node [elem, below of=s12] (s13) {$\sigma_3$};
  \node [right of=s13, node distance=0.75cm] (r1) {};

  \node [elem, right of=s12, node distance=1.5cm] (s22) {$\sigma_2$};
  \node [elem, above of=s22] (s21) {$\sigma_1$};
  \node [elem, below of=s22] (s23) {$\sigma_3$};
  \node [right of=s23, node distance=0.75cm] (r2) {};

  \node [elem, right of=s22, node distance=1.5cm] (s32) {$\sigma_2$};
  \node [elem, above of=s32] (s31) {$\sigma_1$};
  \node [elem, below of=s32] (s33) {$\sigma_3$};
  \node [right of=s33, node distance=0.75cm] (r3) {};

  \node [elem, right of=s32, node distance=1.5cm] (s42) {$\sigma_2$};
  \node [elem, above of=s42] (s41) {$\sigma_1$};
  \node [elem, below of=s42] (s43) {$\sigma_3$};
  \node [right of=s43, node distance=0.75cm] (r4) {};

  \path [line] (s11) -- (s21);
  \path [line] (s21) -- (s32);
  \path [line] (s32) -- (s43);

  \path [rel] (s11) -- (s22);
  \path [rel] (s12) -- (s22);
  \path [rel] (s23) -- (s32);
\end{tikzpicture}
\end{center}
The language contains the following six words
\begin{align*}
(\sigma_1,\sigma_1)(\sigma_1,\sigma_2)(\sigma_2,\sigma_3), \qquad
(\sigma_1,\sigma_2)(\sigma_1,\sigma_2)(\sigma_2,\sigma_3),\\
(\sigma_2,\sigma_2)(\sigma_1,\sigma_2)(\sigma_2,\sigma_3), \qquad
(\sigma_1,\sigma_1)(\sigma_3,\sigma_2)(\sigma_2,\sigma_3),\\
(\sigma_1,\sigma_2)(\sigma_3,\sigma_2)(\sigma_2,\sigma_3), \qquad
(\sigma_2,\sigma_2)(\sigma_3,\sigma_2)(\sigma_2,\sigma_3),
\end{align*}
but only the first,
$(\sigma_1,\sigma_1)(\sigma_1,\sigma_2)(\sigma_2,\sigma_3)$ is
consistent. This word is highlighted with solid arrows in the
diagram above. Now imagine we place the single state pair $(\sigma_2,
\sigma_3)$ into the above language, we might end up with a language as
below:

\begin{center}
\begin{tikzpicture}[x=1cm,auto]
  \node [elem] (s12) {$\sigma_2$};
  \node [elem, above of=s12] (s11) {$\sigma_1$};
  \node [elem, below of=s12] (s13) {$\sigma_3$};
  \node [right of=s13, node distance=0.75cm] (r1) {};

  \node [elem, right of=s12, node distance=1.5cm] (i2) {$\sigma_2$};
  \node [elem, above of=i2] (i1) {$\sigma_1$};
  \node [elem, below of=i2] (i3) {$\sigma_3$};
  \node [right of=i3, node distance=0.75cm] (ri) {};

  \node [elem, right of=i2, node distance=1.5cm] (s22) {$\sigma_2$};
  \node [elem, above of=s22] (s21) {$\sigma_1$};
  \node [elem, below of=s22] (s23) {$\sigma_3$};
  \node [right of=s23, node distance=0.75cm] (r2) {};

  \node [elem, right of=s22, node distance=1.5cm] (s32) {$\sigma_2$};
  \node [elem, above of=s32] (s31) {$\sigma_1$};
  \node [elem, below of=s32] (s33) {$\sigma_3$};
  \node [right of=s33, node distance=0.75cm] (r3) {};

  \node [elem, right of=s32, node distance=1.5cm] (s42) {$\sigma_2$};
  \node [elem, above of=s42] (s41) {$\sigma_1$};
  \node [elem, below of=s42] (s43) {$\sigma_3$};
  \node [right of=s43, node distance=0.75cm] (r4) {};

  \path [rel] (s11) -- (i1);
  \path [rel] (s21) -- (s32);
  \path [line] (s32) -- (s43);
  \path [line] (i2) -- (s23);

  \path [line] (s11) -- (i2);
  \path [line] (s12) -- (i2);
  \path [line] (s23) -- (s32);
\end{tikzpicture}
\end{center}

By performing this shuffle action, we no longer have a consistent word
from $\sigma_1$ to $\sigma_3$, but instead a consistent word from
$\sigma_2$ to $\sigma_3$ and $\sigma_1$ to $\sigma_3$. The new
consistent word was constructed by making a previously inconsistent
word consistent by `filling in the gaps' in that word. This shows that
the shuffle operator can take two inconsistent words and shuffle them
together in such a way that it generates many consistent words. If we
only considered consistent words, we would be unable to define such a
shuffle operator, and would instead have to rely on some operational
semantics to generate traces instead, which would be less
algebraic. This also clearly demonstrates why shuffle is not well
defined on the subalgebra we get by intersecting with $C$.

\newpage

@@
